Entrainment of low Mach number thermals in stratified domains

### 2019-05-20 tex/stratified_thermals.tex
puts sentences onto their own lines for easier future change tracking

\begin{figure}[t!]
    \includegraphics[width=\columnwidth]{overview_fig.png}
    \caption{We have plotted the evolution of the filling factor, or radius
	squared, of a buoyant vortex ring with depth (the $n_\rho = 3$ case examined
	later in this work). Overplotted in a thin
	solid line is the prediction for filling factor growth in the Boussinesq
	case, and in the thin dashed line is plotted the prediction for pure horizontal
	compression, as in \citet{brandenburg2016}.
    \label{fig:overview} }
\end{figure}

@@

\begin{figure}[t!]
    \includegraphics[width=\columnwidth]{evolution_colormeshes.png}
    \caption{The evolution of $\rho S_1 r^3$, or the total entropy weighted by
	the cube of the thermal radius, is shown for two thermals. On the left is
	a thermal in a weakly stratified domain with $n_\rho = 1/2$ density scale heights and 
	on the right is a thermal in a strongly stratified domain with $n_\rho = 3$.
	While both start with precisely the same initial condition, the case with low stratification
	expands with depth like the boussinesq case, whereas the strongly stratified thermal
	compresses with depth.
    \label{fig:evolution_colormeshes} }
\end{figure}

@@

After describing the growth of the thermal's integrated momentum with time, we turn
our attention to th ehydrodynamic impulse, which in the stratified limit is defined
\citep{shivamoggi2010},
\begin{equation}
\bm{I} = \frac{1}{2}\int_{\mathcal{V}} \bm{x}\times(\grad\times(\rho\bm{u}))dV,
\end{equation}
where $\bm{x}$ is the position vector. Impulse is equivalent to the time-integrated
work acting on the fluid resulting in the current fluid motion. 
Per \citet{shivamoggi2010}, changes in the impulse can be expressed as
\begin{equation*}
\frac{D\bm{I}}{D t} = \int_{\mathcal{V}}\frac{\partial(\rho\bm{u})}{\partial t}dV
= B\hat{z} - \int_{\mathcal{V}}\left[\rho\grad\varpi + \Div{\rho\bm{u}\bm{u}}\right]dV ,
\end{equation*}
where, importantly, the \emph{eulerian} derivative of the momentum is inside of the volume
integral here. Under the proper specification of boundary conditions, 
the integral term is zero, and the vertical impulse of a thermal thus straightforwardly
changes in time as
\begin{equation}
\frac{D I_z}{D t} = B.
\label{eqn:change_in_impulse}
\end{equation}
We have thus retrieved the two findings on which our theory of thermal evolution
will be built: \emph{both the impulse and momentum experience constant changes in
time determined by the buoyant nature of the thermal}.

\subsection{Parameterized description of thermal evolution}
\begin{figure}[t!]
    \includegraphics[width=\columnwidth]{constants.png}
    \caption{Plotted are time traces for three quantities which we assume to be constant
	in our thermal theory. In (a) and (b), the circulation and buoyancy, divided by their
	maxima, are plotted vs. depth. We see that with increasing stratification, there is marginally
	more detrainment of both of these quantities, but to first order they are constant over the
	evolution of the thermal. In (c) the constant $f = V/r^3$ is plotted; after significant
	noise during the development of the vortex ring, this quantity remains relatively constant, 
	and increasingly approaches a constant with depth.
    \label{fig:constants} }
\end{figure}

@@

\subsection{Solution for thermal evolution in a Polytrope}
Now that the atmospheric density is specific, we can 
find an explicit solution for Eqn.~\ref{eqn:dz_theory} to compare
to the results of our simulations. As temperature and height are
linearly related in our atmospheres, we find it instructive to use
temperature as a integration parameter.
We note that, $\rho = T^{m_{ad}}$ and
$dT = (\grad_{\text{ad}}) dz_{th}$, such that we must solve
\begin{equation}
\int_{T_0}^{T_{\text{th}}} T^{-m_{ad}/2} dT 
= \left(\frac{\beta(\pi\Gamma)^{3/2}\grad_{\text{ad}}}{f(\chi B)^{1/2}}\right)
\int_{0}^{t + t_{\text{off}}} t^{-1/2} dt.
\end{equation}
defining $C \equiv \beta \pi^{3/2} \grad_{\text{ad}} / f \sqrt{\Gamma^3/(\chi B)}$
and $\alpha^{-1} = 1 - m_{\text{ad}}/2$, and 
ssuming that $m_{\text{ad}} < 2$ (which is valid for our case of $m_{\text{ad}} = 1.5$
studied here) we find that
\begin{equation}
T_{\text{th}} = \left(\frac{2C}{ \alpha } \sqrt{t + t_{\text{off}}} + T_0^{1/\alpha}  \right)^{\alpha}.
\label{eqn:theory_T}
\end{equation}
In the limit of large stratification, we thus find that $T_{\text{th}} \propto t^2$ for
our case of $\alpha = 4$. The temperature can be straightforwardly converted to height
by a rearrangement of Eqn.~\ref{eqn:polytrope},
\begin{equation}
z_{\text{th}} = (T_{\text{th}} - 1)/\grad_{\text{ad}} + L_z
\end{equation}
The thermal is thus described in terms of seven total parameters:
\begin{enumerate}
\item $T_0$ and $t_{\text{off}}$, the temperature of the thermal's virtual origin, and
the time before the simulation's $t = 0$ at which the thermal was located at its virtual origin,
\item $B_0$ and $\Gamma$, the measurable buoyancy and circulation of the vortex ring, and
\item $f$, $\chi$, and $\beta$, fitting parameters which describe the nature of the vortex
ring's evolution.
\end{enumerate}
The measured values of these parameters for the cases studied in this paper are presented
in table \ref{table:parameters}.

@@

\begin{figure}[t!]
    \includegraphics[width=\columnwidth]{results_panels_d_v_t.png}
    \caption{Shown are the measured depths of thermals as a function of time for all 2D
	Anelastic and 3D Fully compressible simulations conducted in this work. Overplotted
	is the theoretical prediction for depth as a function of time.
    \label{fig:results_d_v_t} }
\end{figure}

\begin{figure}[t!]
    \includegraphics[width=\columnwidth]{results_panels_vs_depth.png}
    \caption{Shown are (a) the measured radii of thermals as a function of depth and
	(b) the measured thermal velocities as a function of depth for all 2D
	Anelastic and 3D Fully compressible simulations conducted in this work. Overplotted
	are the predictions from theory.
    \label{fig:results_vs_depth} }
\end{figure}

@@

In Fig.~\ref{fig:results_vs_depth}a, we plot the measured thermal radius vs. depth, with the
theoretical predictions of Eqns.~\ref{eqn:r_theory} \& \ref{eqn:theory_T} plotted as lines
over the data. In the low stratification limit, the radius of the thermal grows linearly with
depth, $r \propto d$, aligning with the Boussinesq limit shown in \citet{lecoanet&jeevanjee2018} 
This growth of the thermal is the result of entrainment of environmental fluid and
results an an accompany decceleration like $w \propto d^{-1}$ in the Boussinesq limit, 
as is shown in Fig.~\ref{fig:results_vs_depth}b.
However, as stratification increases, the thermal entrains less environmental fluid and
expands less, eventually even contracting with depth in the high-stratification limit. 
This lessened entrainment is, unsurprisingly, accompanied by greater acceleration of the thermal.
As in the case of depth vs. time, the overplotted theoretical predictions show excellent
agreement with the measured data.

@@

\subsection{Verification of 2D Anelastic approximation}
\begin{figure*}[t!]
    \includegraphics[width=\textwidth]{diff_AN_FC.png}
    \caption{Measured values of (a) depth vs. time and (c) radius vs. depth are plotted
	for both the 2D anelastic and 3D fully compressible simulations. The fractional
	difference between anelastic and fully compressible results are respectively shown
	in (b) and (d).
    \label{fig:diff} }
\end{figure*}

>>>>

\begin{figure}[t!]
    \includegraphics[width=\columnwidth]{overview_fig.png}
    \caption{
	We have plotted the evolution of the filling factor, or radius squared, of a buoyant vortex ring with depth (the $n_\rho = 3$ case examined later in this work). 
	Overplotted in a thin solid line is the prediction for filling factor growth in the Boussinesq case, and in the thin dashed line is plotted the prediction for pure horizontal compression, as in \citet{brandenburg2016}.
    \label{fig:overview} }
\end{figure}

@@

\begin{figure}[t!]
    \includegraphics[width=\columnwidth]{evolution_colormeshes.png}
    \caption{
	The evolution of $\rho S_1 r^3$, or the total entropy weighted by the cube of the thermal radius, is shown for two thermals. 
	On the left is a thermal in a weakly stratified domain with $n_\rho = 1/2$ density scale heights and on the right is a thermal in a strongly stratified domain with $n_\rho = 3$.
	While both start with precisely the same initial condition, the case with low stratification expands with depth like the boussinesq case, whereas the strongly stratified thermal compresses with depth.
    \label{fig:evolution_colormeshes} }
\end{figure}

@@

After describing the growth of the thermal's integrated momentum with time, we turn our attention to the hydrodynamic impulse, which in the stratified limit is defined
\citep{shivamoggi2010},
\begin{equation}
\bm{I} = \frac{1}{2}\int_{\mathcal{V}} \bm{x}\times(\grad\times(\rho\bm{u}))dV,
\end{equation}
where $\bm{x}$ is the position vector. 
Impulse is equivalent to the time-integrated work acting on the fluid resulting in the current fluid motion. 
Per \citet{shivamoggi2010}, changes in the impulse can be expressed as
\begin{equation*}
\frac{D\bm{I}}{D t} = \int_{\mathcal{V}}\frac{\partial(\rho\bm{u})}{\partial t}dV
= B\hat{z} - \int_{\mathcal{V}}\left[\rho\grad\varpi + \Div{\rho\bm{u}\bm{u}}\right]dV ,
\end{equation*}
where, importantly, the \emph{eulerian} derivative of the momentum is inside of the volume integral here. 
Under the proper specification of boundary conditions, the integral term is zero, and the vertical impulse of a thermal thus straightforwardly changes in time as
\begin{equation}
\frac{D I_z}{D t} = B.
\label{eqn:change_in_impulse}
\end{equation}
We have thus retrieved the two findings on which our theory of thermal evolution will be built: \emph{both the impulse and momentum experience constant changes in time determined by the buoyant nature of the thermal}.

\subsection{Parameterized description of thermal evolution}
\begin{figure}[t!]
    \includegraphics[width=\columnwidth]{constants.png}
    \caption{
	Plotted are time traces for three quantities which we assume to be constant in our thermal theory. 
	In (a) and (b), the circulation and buoyancy, divided by their maxima, are plotted vs. depth. 
	We see that with increasing stratification, there is marginally more detrainment of both of these quantities, but to first order they are constant over the	evolution of the thermal. 
	In (c) the constant $f = V/r^3$ is plotted; after significant noise during the development of the vortex ring, this quantity remains relatively constant, and increasingly approaches a constant with depth.
    \label{fig:constants} }
\end{figure}

@@

\subsection{Solution for thermal evolution in a Polytrope}
Now that the atmospheric density is specific, we can find an explicit solution for Eqn.~\ref{eqn:dz_theory} to compare to the results of our simulations. 
As temperature and height are linearly related in our atmospheres, we find it instructive to use temperature as a integration parameter.
We note that, $\rho = T^{m_{ad}}$ and $dT = (\grad_{\text{ad}}) dz_{th}$, such that we must solve 
\begin{equation}
\int_{T_0}^{T_{\text{th}}} T^{-m_{ad}/2} dT 
= \left(\frac{\beta(\pi\Gamma)^{3/2}\grad_{\text{ad}}}{f(\chi B)^{1/2}}\right)
\int_{0}^{t + t_{\text{off}}} t^{-1/2} dt.
\end{equation}
defining $C \equiv \beta \pi^{3/2} \grad_{\text{ad}} / f \sqrt{\Gamma^3/(\chi B)}$ and $\alpha^{-1} = 1 - m_{\text{ad}}/2$, and assuming that $m_{\text{ad}} < 2$ (which is valid for our case of $m_{\text{ad}} = 1.5$ studied here) we find that
\begin{equation}
T_{\text{th}} = \left(\frac{2C}{ \alpha } \sqrt{t + t_{\text{off}}} + T_0^{1/\alpha}  \right)^{\alpha}.
\label{eqn:theory_T}
\end{equation}
In the limit of large stratification, we thus find that $T_{\text{th}} \propto t^2$ for our case of $\alpha = 4$. 
The temperature can be straightforwardly converted to height by a rearrangement of Eqn.~\ref{eqn:polytrope},
\begin{equation}
z_{\text{th}} = (T_{\text{th}} - 1)/\grad_{\text{ad}} + L_z
\end{equation}
The thermal is thus described in terms of seven total parameters:
\begin{enumerate}
\item $T_0$ and $t_{\text{off}}$, the temperature of the thermal's virtual origin, and the time before the simulation's $t = 0$ at which the thermal was located at its virtual origin,
\item $B_0$ and $\Gamma$, the measurable buoyancy and circulation of the vortex ring, and
\item $f$, $\chi$, and $\beta$, fitting parameters which describe the nature of the vortex ring's evolution.
\end{enumerate}
The measured values of these parameters for the cases studied in this paper are presented in table \ref{table:parameters}.

@@

\begin{figure}[t!]
    \includegraphics[width=\columnwidth]{results_panels_d_v_t.png}
    \caption{
	Shown are the measured depths of thermals as a function of time for all 2D Anelastic and 3D Fully compressible simulations conducted in this work. 
	Overplotted is the theoretical prediction for depth as a function of time.
    \label{fig:results_d_v_t} }
\end{figure}

\begin{figure}[t!]
    \includegraphics[width=\columnwidth]{results_panels_vs_depth.png}
    \caption{
	Shown are (a) the measured radii of thermals as a function of depth and (b) the measured thermal velocities as a function of depth for all 2D Anelastic and 3D Fully compressible simulations conducted in this work. 
	Overplotted are the predictions from theory.
    \label{fig:results_vs_depth} }
\end{figure}

@@

In Fig.~\ref{fig:results_vs_depth}a, we plot the measured thermal radius vs. depth, with the theoretical predictions of Eqns.~\ref{eqn:r_theory} \& \ref{eqn:theory_T} plotted as lines over the data. 
In the low stratification limit, the radius of the thermal grows linearly with depth, $r \propto d$, aligning with the Boussinesq limit shown in \citet{lecoanet&jeevanjee2018}.
This growth of the thermal is the result of entrainment of environmental fluid and results an an accompany decceleration like $w \propto d^{-1}$ in the Boussinesq limit, as is shown in Fig.~\ref{fig:results_vs_depth}b.
However, as stratification increases, the thermal entrains less environmental fluid and expands less, eventually even contracting with depth in the high-stratification limit. 
This lessened entrainment is, unsurprisingly, accompanied by greater acceleration of the thermal.
As in the case of depth vs. time, the overplotted theoretical predictions show excellent agreement with the measured data.

@@

\subsection{Verification of 2D Anelastic approximation}
\begin{figure*}[t!]
    \includegraphics[width=\textwidth]{diff_AN_FC.png}
    \caption{
	Measured values of (a) depth vs. time and (c) radius vs. depth are plotted for both the 2D anelastic and 3D fully compressible simulations. 
	The fractional difference between anelastic and fully compressible results are respectively shown in (b) and (d).
    \label{fig:diff} }
\end{figure*}
